Theoretical aspects in gravity modelling

### 2018-04-27 manuscript/letter.tex
relationship between disturbance and free-air

\begin{summary}
 The gravity anomaly has long been used by geophysicists for 
 representing the gravitational effect produced by gravity
 sources.This quantity, however, can be used in this way as a good
 approximation. Rigorously, the gravitational effect produced by the sources
 is represented by the gravity disturbance, which is a well-known 
 quantity in geodesy.
 In this paper, we discuss the fundamental concepts supporting 
 the fact that gravity disturbance is the more appropriated quantity
 to be used for estimating density variations related to geological
 bodies in subsurface. We stress that almost all authors in the 
 literature have implicitly used the gravity anomaly as an approximation 
 of the gravity disturbance, which can be represented by a harmonic 
 function and, consequently, satisfies the premise behind most of 
 the interpretation techniques used in geophysics (e.g., 
 upward/downward continuation, 
 data processing with equivalent layer,
 conversions between gravity and magnetic data,
 computation of vertical derivatives via Fourier and Hilbert
 transforms).
 Although the gravity anomaly can be considered a good approximation of the 
 gravity disturbance for most practical applications, 
 it is important to bear in mind not only the terminology 
 changes, but also the conceptual assumptions.

@@

\section{Anomalous quantities in gravity modelling}

@@

The gravity anomaly is the commonly used quantity in applied 
geophysics. At point $P$, it is defined as the difference between the 
gravity vector at a point $Q^{\prime}$ on the Geoid (a particular 
equipotential surface of the gravity potential) and the normal gravity 
vector at a point $Q$ on the ellipsoid surface, both at the same
geodetic latitude and longitude (Fig. \ref{fig:surfaces}) as in
\begin{equation}
\Delta \mitbf{g}_{P} = \mitbf{g}_{Q^{\prime} } - \mitbf{\gamma}_{Q} .
\label{eq:gravity-anomaly-vector}
\end{equation}
Similarly as the gravity disturbance expression (eq.
\ref{eq:gravity-disturbance}), the gravity anomaly is given by
\begin{equation}
\Delta g_{P} = g_{Q^{\prime} } - \gamma_{Q} .
\label{eq:gravity-anomaly}
\end{equation}
Notice that, by definition, the gravity anomaly depends on 
longitude and latitude only and is not a function of height 
\citep{barthelmes2013}. As a consequence, it is not possible to compute,
for example, the upward continuation of a gravity anomaly.
However, many authors in the literature have. All these authors have
implicitly considered that the gravity anomaly is an approximation of 
the gravity disturbance, which in turn can be represented by a 
harmonic function that is allowed to be upward continued.

Different gravity anomalies can be calculated (e.g., Free-air and 
Bouguer anomalies), depending on the corrections applied to 
them \citep{blakely1996, hofmann-wellenhof-moritz2005}.
These corrections are usually called gravity reductions.
It can be shown that the Free-air anomaly is an approximation of the
gravity disturbance whereas the Bouguer anomaly
is an approximation of the terrain corrected gravity disturbance.
The Bouguer anomaly is commonly used by geophysicists as the
gravitational effect produced by the gravity sources.
Although this approximation is valid for most practical applications,
it is important to bear in mind not only the terminology 
changes, but also the conceptual assumptions.

@@

The gravity $g_{i}$ can be approximated by a first order Taylor's 
expansion as follows \citep{sanso_sideris2013}:
\begin{equation}
g_{i} \approx \gamma_{i} + 
\hat{\mitbf{\gamma}}_{i}^{\top} \delta \mitbf{g}_{i} \: ,
\label{eq:gobs-approx}
\end{equation}
where $^{\top}$ denotes transposition,
$\hat{\mitbf{\gamma}}_{i} = -\hat{\mitbf{u}}_{i}$ is a unit 
vector with the same direction as the normal gravity vector 
$\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$, in the
topocentric Cartesian coordinate system (Fig. \ref{fig:TCS}b), and
$\hat{\mitbf{u}}_{i}$ is the unit vector $\hat{\mitbf{u}}$ 
(eq. \ref{eq:unit-vectors}) evaluated at the corresponding 
point $(h_{i}, \varphi_{i}, \lambda_{i})$ in the 
GGS (Fig. \ref{fig:GCS-GGS}).
This approximation can be made because, fortunately, 
the condition $\gamma_{i} \gg \| \delta \mathbf{g}_{i} \|$ 
is met at all points located above or on the Earth's surface.
This approximation, which is known in geodesy \citep[e.g.,][]{sanso_sideris2013},
is largely used in applied geophysics for representing
total-field anomalies \citep[e.g.,][]{blakely1996}.
Notice that, in local- or regional-gravity studies, the unit
vector $\hat{\mitbf{\gamma}}_{i}$ (eq. \ref{eq:gobs-approx}) 
may be considered constant throughout the study area and 
parallel to the $z$ axis of the TCS (Fig. \ref{fig:TCS}b).
By using the approximation defined in eq. 
\ref{eq:gobs-approx}, the gravity disturbance 
(eq. \ref{eq:gravity-disturbance}) can be rewritten as 
follows 
\begin{equation}
\delta g_{i} \approx \hat{\mitbf{\gamma}}_{P}^{\top} \delta \mitbf{g}_{i} \: ,
\label{eq:gravity-disturbance-approx}
\end{equation}
where $\hat{\mitbf{\gamma}}_{P}$ represents the unit vector with
the same direction as the normal gravity vector $\mitbf{\gamma}_{P}$
at the origin $P$ of the TCS (Fig. \ref{fig:TCS}b).
This equation shows that the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance}) is different from the magnitude
of the gravity disturbance vector $\delta \mitbf{g}_{i}$
produced by the gravity sources. Rather, it represents the 
component of $\delta \mitbf{g}_{i}$ on the direction of the normal
gravity vector \citep{sanso_sideris2013}.

In the TCS (Fig. \ref{fig:TCS}b),
the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance-approx})
can be defined as the $z$-component of the gravitational attraction exerted by
the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
As a consequence, the gravity disturbance produced by
a gravity source can be represented by the following harmonic function:
\begin{equation}
d_{i} = \partial_{z} \phi_{i} \: ,
\label{eq:gz-local}
\end{equation}
where $\partial_{z} \phi_{i}$ is the first 
derivative, evaluated at the point $(x_{i},y_{i},z_{i})$, of the function 
\begin{equation}
\phi(x,y,z) = \int\int\limits_{v}\int 
\frac{c_{g} \, G \, \Delta\rho(x^{\prime}, y^{\prime}, z^{\prime}) \: dv}
{\sqrt{(x - x^{\prime})^{2} + 
(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: ,
\label{eq:phi}
\end{equation}
with respect to the variable $z$. Is this integral, 
$G$ is the Newtonian constant of gravitation (in $m^{3} kg^{-1} s^{-2}$),
$c_{g} = 10^{5}$ is a constant factor transforming from $m s^{-2}$ 
to milligal (mGal), $\Delta\rho(x^{\prime}, y^{\prime}, z^{\prime})$
is the density contrast (in $kg / m^{3}$) at a point 
$(x^{\prime}, y^{\prime}, z^{\prime})$ within 
the volume $v$ of the gravity source and the integration is conducted 
over $x^{\prime}$, $y^{\prime}$ and $z^{\prime}$.
This equation can be easily generalized for the case
of multiple gravity sources.

@@

\section{Comparing Free-air anomaly with gravity disturbance}

By definition, for a small height $h$, the normal gravity at point
$P(h, \varphi, \lambda)$ can be expanded into a Taylor series 
in terms of $h$ \citep{heiskanen-moritz1967,
hofmann-wellenhof-moritz2005} as in
\begin{equation}
\gamma_{P} \approx \gamma_{Q} + \frac{\partial \gamma}{\partial h} h ,
\label{eq:gamma-expansion}
\end{equation}
where higher orders of $h$ were neglected. $h$ represents the
distance between points $P$ and $Q$ shown in (Fig. \ref{fig:heights}).
Therefore, substituting \ref{eq:gamma-expansion}'s result into
eq. \ref{eq:gravity-disturbance}, the gravity disturbance becomes
\begin{equation}
\delta g_{P} \approx g_{P} - (\gamma_{Q} + \frac{\partial \gamma}{\partial h} h) .
\label{eq:disturb-approx}
\end{equation}
According to \citet{heiskanen-moritz1967, blakely1996}, the relation
between gravity on the geoid and on the reference ellipsoid is
a similar Taylor expansion given by
\begin{equation}
g_{Q^{\prime}} = g_{P} - \frac{\partial \gamma}{\partial h} H
\label{eq:geod2phys_surf}
\end{equation}
where $H$ is defined as the orthometric height (Fig. \ref{fig:heights})
measured along the curved plumb line, starting from the Geoid
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005}.
Thus, the Free-air anomaly $(\Delta g^{a})$, which corrects for the
decrease in gravity due to the elevation of the observation point $P$, 
can be written as
\begin{equation}
\Delta g^{a}_{P} \approx g_{P} - (\gamma_{Q} + \frac{\partial \gamma}{\partial h} H)
\label{eq:free-air}
\end{equation}
By inspection, one easily concludes that the difference is the first
order derivative term
\begin{equation}
\delta g_{P} - \Delta g^{a}_{P} \approx - \frac{\partial \gamma}{\partial h} (h-H) = - \frac{\partial \gamma}{\partial h} N ,
\label{eq:disturb-FA}
\end{equation}
where $N$ is the geoidal undulation (Fig. \ref{fig:heights}) and 
$\frac{\partial \gamma}{\partial h} = -0.3086$.  Hence, using a
mean value of $\tilde{N} = 120$m, the absolute value of the
difference in eq. \ref{eq:disturb-FA} is computed
\begin{equation}
\parallel \delta g_{P} - \Delta g^{a}_{P} \parallel \leq \parallel - \frac{\partial \gamma}{\partial h} \tilde{N} \parallel = 37.032
\label{eq:diff-constant}
\end{equation}

@@

@@

The principal theoretical implication of this study is that,
although the gravity anomaly may be used as a good approximation of
the gravitational effect produced by the sources for most practical applications,
the more appropriated quantity for gravity modelling is the
gravity disturbance. We stress that, more important than the 
the terminology changes, the authors must bear in mind the 
conceptual assumptions used in gravity modelling.

@@

\begin{figure}
    \includegraphics{figures/surfaces.png}
    \caption{Schematic representation of the gravity vector
        $\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$,
        gravity disturbance vector $\delta \mitbf{g}_{P}$
        (eq. \ref{eq:gravity-disturbance-vector}) and unit vector 
        $\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) at a point $P$ on the surface of the Earth; gravity vector 
        $\mitbf{g}_{Q^{\prime}}$ at a point $Q^{\prime}$ on the Geoid;
        normal gravity vector $\mitbf{\gamma}_{Q}$ at a point $Q$ on
        the reference ellipsoid. The dashed line passing through
        $Q$, $Q^{\prime}$ and $P$ is normal to the 
        surface of the reference ellipsoid at $Q$.}
  \label{fig:surfaces}
\end{figure}

\begin{figure}
    \includegraphics{figures/different-heights.png}
    \caption{Schematic representation of the geometric height
        $h$, orthometric height $H$ and geoidal undulation $N$.
        $H$ is curved because it follows the plumbline.}
  \label{fig:heights}
\end{figure}

>>>>

\begin{summary}
 The gravity anomaly has long been used by geophysicists for 
 representing the gravitational effect produced by gravity
 sources.This quantity, however, can be used in this way as a good
 approximation. Rigorously, the gravitational effect produced by the sources
 is represented by the gravity disturbance, which is a well-known 
 quantity in geodesy.
 In this paper, we discuss the fundamental concepts supporting 
 the fact that gravity disturbance is the more appropriated quantity
 to be used for estimating density variations related to geological
 bodies in subsurface. We show that the maximum absolute difference 
 between the gravity disturbance and the Free-air anomaly in Brazil 
 reaches a maximum absolute value of $\approx 9.3$ mGal. We also stress 
 that almost all authors in the literature have implicitly used the 
 gravity anomaly as an approximation of the gravity disturbance, 
 which can be represented by a harmonic function and, consequently, 
 satisfies the premise behind most of 
 the interpretation techniques used in geophysics (e.g., 
 upward/downward continuation, data processing with equivalent layer,
 conversions between gravity and magnetic data, computation of vertical 
 derivatives via Fourier and Hilbert transforms).
 Although the gravity anomaly can be considered a good approximation of the 
 gravity disturbance for most practical applications, 
 it is important to bear in mind not only the terminology 
 changes, but also the conceptual assumptions.

@@

\section{Relationship between gravity disturbance and gravity anomaly}

@@

The gravity anomaly is the commonly used quantity in applied 
geophysics. At point $P$, it is defined as the difference between the 
gravity vector at a point $Q^{\prime}$ on the Geoid (a particular 
equipotential surface of the gravity potential) and the normal gravity 
vector at a point $Q$ on the ellipsoid surface, both at the same
geodetic latitude and longitude (Fig. \ref{fig:surfaces}) as in
\begin{equation}
\Delta \mitbf{g}_{P} = \mitbf{g}_{Q^{\prime}} - \mitbf{\gamma}_{Q} .
\label{eq:gravity-anomaly-vector}
\end{equation}
Similarly to the gravity disturbance expression (eq.
\ref{eq:gravity-disturbance}), the gravity anomaly is given by
\begin{equation}
\Delta g_{P} = g_{Q^{\prime}} - \gamma_{Q} ,
\label{eq:gravity-anomaly}
\end{equation}
where $g_{Q^{\prime} } = \| \mitbf{g}_{Q^{\prime}} \|$ is
the magnitude of the gravity vector $\mitbf{g}_{Q}$ at $Q^{\prime}$, 
on the Geoid, and $\gamma_{Q} = \| \mitbf{\gamma}_{Q} \|$ is the 
magnitude of the normal gravity vector $\mitbf{\gamma}_{Q}$ at $Q$, 
on the reference ellipsoid (Fig. \ref{fig:surfaces}).
Notice that, by definition, the gravity anomaly depends on 
longitude and latitude only and is not a function of height 
\citep{barthelmes2013}. As a consequence, it is not possible to compute,
for example, the upward continuation of a gravity anomaly.
However, many authors in the literature have computed the upward 
continuation of gravity anomalies. All these authors have
implicitly considered that the gravity anomaly is an approximation of 
the gravity disturbance, which in turn can be represented by a 
harmonic function that is allowed to be upward continued.

Different gravity anomalies can be calculated (e.g., Free-air and 
Bouguer anomalies), depending on the corrections applied to 
them. These corrections are usually called gravity reductions. 
The Free-air anomaly, for example, is defined as follows 
\citep{blakely1996, hofmann-wellenhof-moritz2005}:
\begin{equation}
\Delta g_{P}^{a} 
= g_{P} - \left( \gamma_{Q} + \frac{\partial \gamma}{\partial h} \, H \right)
\: ,
\label{eq:free-air-anomaly}
\end{equation}
where $\frac{\partial \gamma}{\partial h} \approx -0.3086$ mGal/m is the
derivative of the normal gravity with respect to the geometric height $h$
and $H$ is the orthometric height (Fig. \ref{fig:surfaces}). 
Notice that the term between parenthesis represents an 
approximation of the normal gravity at the point $Q^{\prime}$ on the
Geoid. By using a similar approach, we can approximate the normal gravity
$\gamma_{P}$, at a point $P$ on the Earth's surface, so that the 
gravity disturbance $\delta g_{P}$ (eq. \ref{eq:gravity-disturbance}) 
can be rewritten as follows:
\begin{equation}
\delta g_{P} \approx g_{P} - 
\left( \gamma_{Q} + \frac{\partial \gamma}{\partial h} h \right) \: .
\label{eq:gravity-disturbance-approx}
\end{equation}
By inspection, one easily concludes that the absolute difference 
between the approximated gravity disturbance $\delta g_{P}$ 
(eq. \ref{eq:gravity-disturbance-approx}), at a point $P$ on or close to
the Earth's surface, and the Free-air anomaly
$\Delta g_{P}^{a}$ (eq. \ref{eq:free-air-anomaly}), is given by:
\begin{equation}
\vert \delta g_{P} - \Delta g^{a}_{P} \vert \approx 
\vert \frac{\partial \gamma}{\partial h} N \vert \: ,
\label{eq:difference-disturbance-free-air}
\end{equation}
where the geoidal undulation $N \approx h - H$ (Fig. \ref{fig:surfaces}) assumes 
that the Geoid and the surface of the reference ellipsoid 
are approximately parallel at $P$ and the surrounding area.
We know empirically that the geoidal undulation $N$ in the world 
is $\approx \pm 1$ on the oceans and reaches a 
maximum absolute value of $\approx 120$ m \citep[e.g.,][]{torge2012, 
sanso_sideris2013}. In Brazil, for example, the geoidal undulation 
reaches $\approx \pm 30$ m \citep{ibge_mapgeo2015}. Consequently, the
maximum absolute differences between gravity disturbance and
Free-air anomaly, according to eq. \ref{eq:difference-disturbance-free-air},
is $\approx 9.258$ mGal.

The approximations defined by eqs \ref{eq:gravity-disturbance-approx}
and \ref{eq:difference-disturbance-free-air} are commonly used
in geodesy to define the \textit{fundamental equation of physical
geodesy} \citep{hofmann-wellenhof-moritz2005}.
A similar approach can be used to show that the Bouguer anomaly
is an approximation of the terrain corrected gravity disturbance.
The Bouguer anomaly is commonly used by geophysicists as the
gravitational effect produced by the gravity sources.
Although this approximation is valid for most practical applications,
it is important to bear in mind not only the terminology 
changes, but also the conceptual assumptions.

@@

The gravity $g_{i}$ can be approximated by a first order Taylor's 
expansion as follows \citep{sanso_sideris2013}:
\begin{equation}
g_{i} \approx \gamma_{i} + 
\hat{\mitbf{\gamma}}_{i}^{\top} \delta \mitbf{g}_{i} \: ,
\label{eq:gobs-approx}
\end{equation}
where $^{\top}$ denotes transposition,
$\hat{\mitbf{\gamma}}_{i} = -\hat{\mitbf{u}}_{i}$ is a unit 
vector with the same direction as the normal gravity vector 
$\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$, in the
topocentric Cartesian coordinate system (Fig. \ref{fig:TCS}b), and
$\hat{\mitbf{u}}_{i}$ is the unit vector $\hat{\mitbf{u}}$ 
(eq. \ref{eq:unit-vectors}) evaluated at the corresponding 
point $(h_{i}, \varphi_{i}, \lambda_{i})$ in the 
GGS (Fig. \ref{fig:GCS-GGS}).
This approximation can be made because, fortunately, 
the condition $\gamma_{i} \gg \| \delta \mathbf{g}_{i} \|$ 
is met at all points located above or on the Earth's surface.
This approximation, which is known in geodesy \citep[e.g.,][]{sanso_sideris2013},
is largely used in applied geophysics for representing
total-field anomalies \citep[e.g.,][]{blakely1996}.
Notice that, in local- or regional-gravity studies, the unit
vector $\hat{\mitbf{\gamma}}_{i}$ (eq. \ref{eq:gobs-approx}) 
may be considered constant throughout the study area and 
parallel to the $z$ axis of the TCS (Fig. \ref{fig:TCS}b).
By using the approximation defined in eq. 
\ref{eq:gobs-approx}, the gravity disturbance 
(eq. \ref{eq:gravity-disturbance}) can be rewritten as 
follows 
\begin{equation}
\delta g_{i} \approx \hat{\mitbf{\gamma}}_{P}^{\top} \delta \mitbf{g}_{i} \: ,
\label{eq:gravity-disturbance-approx-TCS}
\end{equation}
where $\hat{\mitbf{\gamma}}_{P}$ represents the unit vector with
the same direction as the normal gravity vector $\mitbf{\gamma}_{P}$
at the origin $P$ of the TCS (Fig. \ref{fig:TCS}b).
This equation shows that the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance}) is different from the magnitude
of the gravity disturbance vector $\delta \mitbf{g}_{i}$
produced by the gravity sources. Rather, it represents the 
component of $\delta \mitbf{g}_{i}$ on the direction of the normal
gravity vector \citep{hofmann-wellenhof-moritz2005, sanso_sideris2013}.

In the TCS (Fig. \ref{fig:TCS}b),
the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance-approx-TCS})
can be defined as the $z$-component of the gravitational attraction exerted by
the gravity sources at the point $(x_{i}, y_{i}, z_{i})$.
As a consequence, the gravity disturbance produced by
a gravity source can be represented by the following harmonic function:
\begin{equation}
d_{i} = \partial_{z} \phi_{i} \: ,
\label{eq:gz-local}
\end{equation}
where $\partial_{z} \phi_{i}$ is the first 
derivative, evaluated at the point $(x_{i},y_{i},z_{i})$, of the function 
\begin{equation}
\phi(x,y,z) = \int\int\limits_{v}\int 
\frac{c_{g} \, G \, \Delta\rho(x^{\prime}, y^{\prime}, z^{\prime}) \: dv}
{\sqrt{(x - x^{\prime})^{2} + 
(y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: ,
\label{eq:phi}
\end{equation}
with respect to the variable $z$. Is this integral, 
$G$ is the Newtonian constant of gravitation (in $m^{3} kg^{-1} s^{-2}$),
$c_{g} = 10^{5}$ is a constant factor transforming from $m s^{-2}$ 
to milligal (mGal), $\Delta\rho(x^{\prime}, y^{\prime}, z^{\prime})$
is the density contrast (in $kg / m^{3}$) at a point 
$(x^{\prime}, y^{\prime}, z^{\prime})$ within 
the volume $v$ of the gravity source and the integration is conducted 
over $x^{\prime}$, $y^{\prime}$ and $z^{\prime}$.
This equation can be easily generalized for the case
of multiple gravity sources.

@@

@@

\item In Brazil, the maximum difference between the gravity 
disturbance (on or close to the Earth's surface) and the Free-air
anomaly reaches $\approx 9.3$ mGal.

@@

The principal theoretical implication of this study is that,
although the gravity anomaly may be used as a good approximation of
the gravitational effect produced by the sources for most practical applications,
the more appropriated quantity for gravity modelling is the
gravity disturbance. We stress that, more important than the 
terminology changes, the researchers must bear in mind the 
conceptual assumptions used in gravity modelling.

@@

\begin{figure}
    \includegraphics{figures/surfaces.png}
    \caption{Schematic representation of the gravity vector
        $\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$,
        gravity disturbance vector $\delta \mitbf{g}_{P}$
        (eq. \ref{eq:gravity-disturbance-vector}), unit vector 
        $\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) at a point $P$ 
        on the surface of the Earth, gravity vector $\mitbf{g}_{Q^{\prime}}$ 
        at a point $Q^{\prime}$ on the Geoid, normal gravity vector 
        $\mitbf{\gamma}_{Q}$ at a point $Q$ on the reference ellipsoid,
        geometric height $h$, orthometric height $H$ and geoidal undulation 
        $N$ \citep{heiskanen-moritz1967}.
        The dashed line passing through
        $Q$, $Q^{\prime}$ and $P$ is normal to the 
        surface of the reference ellipsoid at $Q$. This figure
        shows a commonly used approximation in which the ellipsoid 
        surface and the Geoid are represented as parallel surfaces, 
        so that $h \approx H + N$.}
  \label{fig:surfaces}
\end{figure}
